Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-03 paper/metamonads.tex
Finish comments on modularization design. Show DOMAIN and CONTAINER in full.

A careful study of Gaussian Elimination \cite{carette04} shows that
(at least) the following variations occur:
\begin{enumerate}
	\item \textbf{Domain}: In which (algebraic) domain do the matrix
		elements belong to.  Some implementations were very specific
		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
		example), while others were generic for elements of a field,
		multivariate polynomials over a field, elements of a formal Ring
		with possibly undecidable zero-equivalence, or elements of a 
		Division Ring.  In the roughly 85 pieces of code surveyed,
		20 different domains were encountered.
	\item \textbf{Container}: Whether the matrix
		was represented as an array of arrays, a one-dimensional array,
		a hash table, a sparse matrix, etc., and
		whether indexing was done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this was sometimes used.  
	\item \textbf{Output choices}:  Whether just the reduced matrix, or
		additionally the rank, the determinant, and the
		pivoting matrix were to be returned.
		It is worthwhile noting that in the larger algorithm family,
		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
		algorithm is allowed to use unrestricted division, or only
		exact (remainder-free) division.
	\item \textbf{Pivoting}: Whether to use no pivoting at all, only
		column-wise pivoting, or full pivoting.
	\item \textbf{Augmented Matrices}: Whether we are doing GE on
		a full matrix, or only a restricted number of columns, while
		doing elimination on the full matrix.  Note that we currently
		do not treat this aspect in our code.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\begin{enumerate}
	\item \textbf{Length measure}:  For stability reasons
		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this was sometimes used to choose
		an ``optimal'' pivot for row exchanges.
	\item \textbf{Normalization and zero-equivalence}: Whether the 
		arithmetic operations of the domain at hand gives results in 
		normalized form, and whether a specialized zero-equivalence 
		routine needs to be used.
\end{enumerate}
\noindent These are separated out from the others are they are cross-cutting
concerns: in the case of the length measure, then a property of the domain
will influence the method used to do pivoting \textbf{if} pivoting is to be
performed.

Simple parameterization: by the domain, by the container

>>>>

A careful study of Gaussian Elimination \cite{carette04} shows that
(at least) the following variations occur:
\begin{enumerate}
	\item \textbf{Domain}: In which (algebraic) domain do the matrix
		elements belong to.  Sometimes the domains are very specific
		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p, 
		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
		example), while in other cases the domains were left generic,
		namely for elements of a field,
		multivariate polynomials over a field, elements of a formal Ring
		with possibly undecidable zero-equivalence, or elements of a 
		Division Ring.  In the roughly 85 pieces of code surveyed,
		20 different domains were encountered.
	\item \textbf{Container}: Whether the matrix
		is represented as an array of arrays, a one-dimensional array,
		a hash table, a sparse matrix, etc., and
		whether indexing is done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this is sometimes used.  
	\item \textbf{Output choices}:  Whether just the reduced matrix, or
		additionally the rank, the determinant, and the
		pivoting matrix are to be returned.
		It is worthwhile noting that in the larger algorithm family,
		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
		algorithm is allowed to use unrestricted division, or only
		exact (remainder-free) division.
	\item \textbf{Pivoting}: Whether to use no pivoting at all, only
		column-wise pivoting, or full pivoting.
	\item \textbf{Augmented Matrices}: Whether we are doing GE on
		a full matrix, or only a restricted number of columns, while
		doing elimination on the full matrix.  Note that we currently
		do not treat this aspect in our code.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\begin{enumerate}
	\item \textbf{Length measure}:  For stability reasons
		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this is sometimes used to choose
		an ``optimal'' pivot for row exchanges.
	\item \textbf{Normalization and zero-equivalence}: Whether the 
		arithmetic operations of the domain at hand gives results in 
		normalized form, and whether a specialized zero-equivalence 
		routine needs to be used.
\end{enumerate}
\noindent These are separated out from the others are they are cross-cutting
concerns: in the case of the length measure, then a property of the domain
will influence the method used to do pivoting \textbf{if} pivoting is to be
performed.

The simplest parametrization is to make the domain abstract.  As it 
turns out, we need the following to exist in our domains:  $0$, $1$,
$+$, $*$, (unary and binary) $-$, some kind of division, normalization,
and potentially a relative size measure.  We definitely need at least
\emph{exact} division to exist, but if full division is available, than
can be used as well.  Normalization is frequently trivial for domains
where elements have unique representations, but in many applications
we have domains (like polynomials in sum-of-product form) where this
is not the case.  This leads us to defining a |DOMAIN| type as
\begin{small}
\begin{code}
module type DOMAIN = sig
  type v
  type kind (* Field or Ring ? *)
  type 'a vc = ('a,v) code
  val zero : 'a vc
  val one : 'a vc
  val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
  val times : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
  val minus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
  val uminus : 'a vc -> ('a vc, 's, 'w) monad
  val div : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
  val better_than : ('a vc -> 'a vc -> 
      (('a,bool) code, 's, 'w) monad) option
  val normalizerf : (('a,v -> v) code ) option
  val normalizerg : 'a vc -> 'a vc
end 
\end{code}
\end{small}
\noindent  The first item to notice is that the types above are
generally lifted twice: once from the value domain |v| to the code
domain |'a vc|, and once more from values to monadic computations
|('a vc, 's, 'w) monad|.  Since there are times where we don't want
to redundantly walk a whole matrix to apply normalization when 
normalization is not necessary, |normalizerf| is optional.

For example, when lifted to a |DOMAIN|, the type |int| becomes
\begin{small}
\begin{code}
module IntegerDomain = struct
    type v = int
    type kind = domain_is_ring
    type 'a vc = ('a,v) code
    let zero = .< 0 >.  
    let one = .< 1 >. 
    let plus x y = ret .<.~x + .~y>. 
    let times x y = ret .<.~x * .~y>.
    let minus x y = ret .<.~x - .~y>.
    let uminus x = ret .< - .~x>.
    let div x y = ret .<.~x / .~y>. 
    let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
    let normalizerf = None 
    let normalizerg = fun x -> x
end
\end{code}
\end{small}
\noindent While the |DOMAIN| type may have looked daunting to some, 
the implementation is quite straightforward.  It is equally simple to 
make implementations for |float| and for |Num.num| (arbitrary precision
exact rational numbers).  With scarcely more work (as these are not 
built-in and the arithmetic must be implemented), $\mathbb{Z}_p$,
$\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$ and more
complex domains can be done as well.

Parametrizing by the kind of container used to represent a matrix is
almost as straightforward.  We choose to make the container parametric
over a |DOMAIN| by making containers functors from a |DOMAIN| module
to the actual implementation of a container.  Our containers are
explicitly 2-dimensional, and we provide functions |dim1| and |dim2|
to extract each.  We also need |get| and |set| functions -- these come
in both a monadic flavour as well as a direct code-generation version.
For convenience and efficiency, several additional functions are also
required: a way to |copy| a container, a method (|mapper|) to map an
optional function over the elements of a container, as well as methods
to swap rows and columns.  The inclusion of these functions in the 
signature of all containers makes it simpler to optimize the relevant
functions depending on the actual representation of the container
while not burdening the users of containers with efficiency details.
As an example, below is an implementation of a matrix as an
array of rows (themselves arrays).  This implementation clearly shows
how swapping rows is more efficient than swapping columns in this
representation.  We have not (yet) implemented the further potential
optimization wherein if it is known that part of two rows/columns are already
known to be equal (in fact to |Dom.zero| in the case of GE) and
row/column exchange needs to be performed element-wise, then certain
swaps are unecessary.

\begin{small}
\begin{code}
module GenericArrayContainer(Dom:DOMAIN) =
  struct
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let get' x n m = .< (.~x).(.~n).(.~m) >.
  let get x n m = ret (get' x n m)
  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
  let set x n m y = ret (set' x n m y)
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  let mapper g a = match g with
      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
      | None   -> a
  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
                       (Array.copy .~a) >. )
  (* this can be optimized with a swap_rows_from if it is known that
     everything before that is already = Dom.zero *)
  let swap_rows_stmt a r1 r2 =
      .< let t = (.~a).(.~r1) in
         begin 
             (.~a).(.~r1) <- (.~a).(.~r2);
             (.~a).(.~r2) <- t
         end >.
  let swap_cols_stmt a c1 c2 = .< 
      for r = 0 to .~(dim2 a)-1 do
          let t = (.~a).(r).(.~c1) in
          begin 
              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
              (.~a).(r).(.~c2) <- t
          end
      done  >.
end
\end{code}
\end{small}
\noindent  Implementing other containers based on flattened vectors,
Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
Naturally, for proper use of |Bigarray| with specific types, it is 
important that the |DOMAIN| type matches -- but luckily the compiler 
will prevent us from making such a mistake.

The use of a |functor| for making a container parametric is fairly 
straightforward.  It gets more interesting when one turns to the aspect
of what to return from the GE algorithm.
